---
title: "diag_test() & roc(): Diagnostic Accuracy and Biomarker Discrimination"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{diag_test() & roc(): Diagnostic Accuracy and Biomarker Discrimination}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
library(SimtablR)
data(epitabl)
has_pROC <- requireNamespace("pROC", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

Diagnostic evaluations in clinical epidemiology assess how accurately an index test identifies the presence or absence of a target condition compared against a gold-standard reference.

In this guide, we follow the acute diagnostic phase of the **ESTROBE-ACS study**. Across six emergency departments, patients presenting with suspected acute coronary syndrome (ACS) were enrolled in a point-of-care (POC) biomarker substudy. We evaluate whether qualitative bedside point-of-care high-sensitivity troponin (`poc_hstn_positive`) or quantitative troponin concentrations (`poc_hstn_value`) can accurately rule out or identify adjudicated ACS (`adjudicated_acs`), adhering to the **STARD 2015** (Standards for Reporting Diagnostic Accuracy Studies) guidelines.

---

## 1. Defining the Diagnostic Cohort

> **Clinical Context & Research Question:** Within the ESTROBE-ACS emergency cohort, how was the point-of-care biomarker substudy sampled, and what are the observed test completion counts?

Before evaluating test performance, clinical investigators must document participant accounting:

```{r diagnostic-sample}
# Subset participants enrolled in the bedside biomarker substudy
diagnostic <- subset(epitabl, diagnostic_substudy == "Yes")

c(
  total_enrolled = nrow(epitabl),
  substudy_enrolled = nrow(diagnostic),
  observed_index_tests = sum(!is.na(diagnostic$poc_hstn_positive)),
  observed_reference_standards = sum(!is.na(diagnostic$adjudicated_acs))
)
```

---

## 2. Binary Diagnostic Test Accuracy with `diag_test()`

> What is the clinical accuracy (Sensitivity, Specificity, PPV, NPV, and Likelihood Ratios) of a qualitative bedside troponin test for detecting adjudicated ACS?

`diag_test()` evaluates binary index tests against a reference standard. To ensure clinical transparency, you must explicitly declare which factor levels represent positive status:
- `positive`: The target condition level in the reference standard (`"Yes"`).
- `test_positive`: The positive result level in the index test (`"Positive"`).

```{r diag-accuracy}
accuracy <- diag_test(
  diagnostic,
  test = poc_hstn_positive,
  ref = adjudicated_acs,
  positive = "Yes",
  test_positive = "Positive",
  ci = "exact"
)
accuracy
```

### Key Metrics Computed:
- **Sensitivity & Specificity:** Probability of a positive test given true disease, and negative test given absence of disease.
- **PPV & NPV (Predictive Values):** Positive and Negative predictive values calculated at the cohort prevalence.
- **Positive & Negative Likelihood Ratios ($\text{LR}^+, \text{LR}^-$):** Prevalence-independent measures of diagnostic revision.
- **Diagnostic Odds Ratio (DOR):** Ratio of odds of positive test in diseased vs non-diseased ($\frac{\text{TP} \times \text{TN}}{\text{FP} \times \text{FN}}$).
- **Confidence Intervals (`ci`):** Supports exact Clopper-Pearson binomial intervals (`"exact"`) or score-based Wilson intervals (`"wilson"`).

---

### 2.1 Confusion Matrix Visualization

SimtablR provides a fourfold display of the confusion matrix via `plot()`:

```{r diag-plot, fig.alt="Fourfold plot showing true positives, false positives, false negatives, and true negatives"}
plot(accuracy, main = "Point-of-Care Troponin Confusion Matrix")
```

The four quadrants represent True Positives, False Positives, False Negatives, and True Negatives, visually scaling cell counts and odds ratios.

---

### 2.2 Tidy Metric Extraction

All underlying metrics, standard errors, and confidence intervals are stored without destructive rounding in `$data` and can be extracted as a tidy data frame:

```{r diag-extract}
# Extract structured evidence table
head(as.data.frame(accuracy, tidy = TRUE), 8)
```

---

## 3. Continuous Biomarker Discrimination with `roc()`

> How well does quantitative bedside troponin concentration discriminate adjudicated ACS across the entire spectrum of decision cutoffs, and what is the optimal threshold by Youden's J index?

When an index marker is measured on a continuous numerical scale (such as `poc_hstn_value` in ng/L), `roc()` computes the Receiver Operating Characteristic curve.

### 3.1 Empirical AUC and DeLong Confidence Intervals

Evaluating the marker without cutpoint hunting provides an unbiased assessment of overall discriminatory capacity:

```{r roc-curve, eval=has_pROC}
troponin_roc <- roc(
  diagnostic,
  marker = poc_hstn_value,
  outcome = adjudicated_acs,
  positive = "Yes",
  cutpoint = "none"
)
troponin_roc
```

The estimated Area Under the Curve (AUC) is reported alongside asymptotic DeLong 95% confidence intervals.

---

### 3.2 Threshold Optimization with Youden's J

In clinical practice, emergency physicians often require a single decision cutoff to trigger admission or discharge. Passing `cutpoint = "youden"` optimizes the trade-off between Sensitivity and Specificity by maximizing Youden's $J = \text{Sensitivity} + \text{Specificity} - 1$:

```{r roc-youden, eval=has_pROC}
troponin_youden <- roc(
  diagnostic,
  marker = poc_hstn_value,
  outcome = adjudicated_acs,
  positive = "Yes",
  cutpoint = "youden"
)
troponin_youden
```

SimtablR automatically displays educational advice reminding analysts that data-derived thresholds exhibit optimistic bias and require external cohort validation before clinical deployment.

---

### 3.3 Publication ROC Plot

SimtablR includes an `autoplot()` method that produces a publication-ready ROC curve:

```{r roc-plot, eval=has_pROC && has_ggplot2, fig.alt="ROC curve displaying sensitivity versus 1 - specificity"}
ggplot2::autoplot(troponin_roc) +
  ggplot2::labs(
    title = "ROC Discrimination: Point-of-Care Troponin",
    subtitle = "ESTROBE-ACS Diagnostic Substudy"
  )
```

---

## 4. Reporting Guidelines: STARD Checklist & Automated Methods

> How do we verify methodological compliance against the STARD 2015 diagnostic checklist and generate publication-ready methods text?

SimtablR facilitates guideline adherence through the `stard()` reporting verb:

```{r stard-check}
stard_report <- stard(accuracy)
head(as.data.frame(stard_report), 8)
```

Furthermore, `as_methods()` translates the exact analytical choices, confidence interval formulas, and software implementations into concise prose ready for manuscript submission:

```{r stard-methods}
cat(as_methods(accuracy))
```
