---
title: "simtab Grammar, Reviewer Analyses, and Export: From Protocol to Manuscript"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simtab Grammar, Reviewer Analyses, and Export: From Protocol to Manuscript}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
library(SimtablR)
data(epitabl)
has_gt <- requireNamespace("gt", quietly = TRUE)
has_flextable <- requireNamespace("flextable", quietly = TRUE)
has_office <- has_flextable &&
  requireNamespace("officer", quietly = TRUE) &&
  requireNamespace("openxlsx", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

In academic and clinical epidemiology, producing an analysis is only the beginning: investigators must document analysis plans, respond to peer-review critique with sensitivity and bias analyses, reconcile participant flow, and format manuscript-ready tables for scientific journals.

In this guide, we finalize the **ESTROBE-ACS study** for publication. We demonstrate how SimtablR's declarative grammar (`simtab`), composite reporting (`simtablr`), reviewer toolkit (`sensitivity`, `e_value`, `flow`), methodological auditor (`advise`), and multi-format export engines transition research from protocol to publication.

---

## 1. The Declarative SimtablR Grammar

> **Clinical Context & Research Question:** In pre-registered study protocols, how do we define an inert, audit-ready analysis specification before executing calculations, and how does verb closure allow post-hoc refinements?

While direct functions like `table1()` and `tb()` provide quick entry points, the declarative grammar allows you to define, inspect, and modify an analysis plan explicitly before computation.

### 1.1 Building an Inert Specification (`simtab_spec`)

Calling `simtab()` records data references, variables, summary policies, and formatting into an inert `simtab_spec` object without executing statistical tests:

```{r grammar-spec}
plan <- simtab(epitabl) |>
  describe(c(age, sex, bmi, renal_impairment)) |>
  stratify(adjudicated_acs) |>
  set_summary("auto") |>
  missingness(display = TRUE) |>
  overall(TRUE) |>
  test("auto") |>
  label(
    age = "Age (years)",
    bmi = "Body mass index (kg/m²)"
  ) |>
  fmt(d = 1)

# Printing displays the declared plan without computing
plan
```

### 1.2 Plan Validation & Evaluation

Before running calculations, `validate()` confirms the plan is internally consistent. Then `evaluate()` computes the numerical evidence:

```{r grammar-eval}
validate(plan)

baseline_result <- evaluate(plan)
baseline_result
```

### 1.3 Verb Closure: Modifying Existing Results

A core design feature of SimtablR is **verb closure**: grammar verbs can be piped directly into computed `simtab_result` objects. SimtablR updates the stored specification and recomputes without mutating the original object:

```{r verb-closure}
# Add Standardized Mean Differences (SMD) to the computed baseline table
baseline_smd <- baseline_result |> test(smd = TRUE)
baseline_smd
```

---

## 2. Composite Manuscript Reports with `simtablr()`

> How do we assemble a complete epidemiological study report—pairing a descriptive Table 1 (demographics) with an inferential Table 2 (crude and adjusted associations)—in a single call?

Scientific manuscripts standardly report baseline characteristics (Table 1) alongside multivariable associations for the primary exposure (Table 2). `simtablr()` builds this combined report in one call:

```{r simtablr-report}
report <- simtablr(
  epitabl,
  outcome = adjudicated_acs,
  exposure = renal_impairment,
  vars = c(age, sex, smoking, hypertension, diabetes, renal_impairment),
  adjust = c(age, sex, smoking, hypertension),
  design = "cohort"
)
report
```

You can inspect individual components directly:
- `report$table1`: Baseline descriptive table.
- `report$table2`: Bivariate and multivariable association table.

---

## 3. Reviewer Analyses, Sensitivity, and Bias Auditing

> When responding to journal peer reviewers, how do we evaluate sensitivity to alternative scales, quantify resilience to unmeasured confounding via E-values, and account for patient attrition?

During peer review, reviewers frequently request sensitivity tests, bias assessments, and methodological justifications.

### 3.1 Sensitivity Analysis with `sensitivity()`

`sensitivity()` tests how results behave under alternative modeling decisions without overwriting the primary analysis:

```{r sensitivity-demo}
# Primary association
primary <- tb(
  epitabl,
  renal_impairment,
  adjudicated_acs,
  flags = c("row", "rr"),
  design = "cohort"
)

# Test sensitivity to odds ratio scale and complete-case denominator
sens_res <- sensitivity(
  primary,
  measure = "OR",
  denominator = "complete"
)
sens_res
```

### 3.2 Unmeasured Confounding: E-Value Analysis

To evaluate vulnerability to unmeasured confounding, `e_value()` computes the minimum association strength that an unmeasured confounder must have with both the exposure and outcome to explain away the observed estimate:

```{r evalue-demo}
ev <- e_value(primary)
ev
```

### 3.3 Methodological Advice and Decision Audit

SimtablR's advice engine audits analyses against methodological best practices:

```{r advise-audit}
# Review fired advice entries
advise(primary)

# Conduct an exhaustive audit of all rules (fired and silent)
audit_report <- advise(primary, audit = TRUE)
audit_report
```

To explain the internal statistical decisions and fallback rules applied during computation:

```{r why-decisions}
why(primary)
```

### 3.4 Participant Flow and Flowchart Visualization

Reconcile initial participant enrollment against the analyzed sample using `flow()`:

```{r flow-demo}
study_flow <- flow(report)
study_flow
```

```{r flow-plot, eval=has_ggplot2, fig.alt="Participant flowchart showing study attrition"}
ggplot2::autoplot(study_flow) +
  ggplot2::labs(title = "Participant Flow: ESTROBE-ACS Cohort")
```

### 3.5 Reporting Checklists and Codebooks

To verify reporting compliance against international standards (STROBE for observational studies; STARD for diagnostic accuracy):

```{r strobe-demo}
strobe_summary <- strobe(report)
head(as.data.frame(strobe_summary), 6)
```

Generate a structured data dictionary / codebook:

```{r codebook-demo}
cb <- codebook(epitabl[, c("age", "sex", "hypertension", "adjudicated_acs")])
head(cb)
```

---

## 4. Publication Rendering, Journal Styling, and File Export

> How do we apply target journal formatting (e.g., JAMA, NEJM) and export final tables directly to Microsoft Word (.docx), Excel (.xlsx), and interactive HTML?

SimtablR strictly separates statistical computation from journal presentation. Tables can be styled according to target journal guidelines and exported natively to Office formats.

### 4.1 Journal Presets

Apply journal formatting presets via `style()`:

```{r journal-style}
# List registered journal presets
list_journals()

# Apply JAMA styling (formatting, confidence intervals, p-value thresholds)
jama_table <- style(primary, "jama")
jama_table
```

### 4.2 Interactive Tables (`gt`) and Office Tables (`flextable`)

SimtablR outputs directly to `gt` for HTML documents and `flextable` for Microsoft Word / PowerPoint:

```{r table-rendering, eval=has_gt}
as_gt(report$table2)
```

```{r flextable-rendering, eval=has_flextable}
flextable::as_flextable(report$table2)
```

### 4.3 Native Export to Word and Excel

Export complete tables and composite reports directly to `.docx` or `.xlsx` files:

```{r export-demo, eval=has_office}
tmp_docx <- tempfile(fileext = ".docx")
tmp_xlsx <- tempfile(fileext = ".xlsx")

export_docx(report, path = tmp_docx)
export_xlsx(report, path = tmp_xlsx)

# Verify exported files exist
file.exists(tmp_docx)
file.exists(tmp_xlsx)

# Clean up temporary demonstration files
unlink(c(tmp_docx, tmp_xlsx))
```

### 4.4 Automated Methodology Prose

Finally, SimtablR automatically composes a reproducible Methods section describing the analytical steps taken:

```{r methods-prose}
cat(as_methods(report))
```
